---
title: "Getting Started with LDNDC in PEcAn"
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  warning: true
  message: true
---

## Purpose

This tutorial is for LDNDC users who have already completed Demo 1 in
`documentation/tutorials/Demo_1_Basic_Run/` and now want to run an LDNDC example
through the PEcAn model-coupler workflow.

- LDNDC is a model coupled to PEcAn; the binary is downloaded from the
  [LandscapeDNDC binary distributions](https://ldndc.imk-ifu.kit.edu/download/download-model.php)
  page.
- `PEcAn.LDNDC::met2model.LDNDC()` converts PEcAn meteorology into LDNDC climate input.
- `PEcAn.LDNDC::write.config.LDNDC()` writes the LDNDC project files and `job.sh`.
- `PEcAn.LDNDC::model2netcdf.LDNDC()` converts raw LDNDC output to PEcAn-standard yearly NetCDF.

The worked example is based on the grassland/meadow example provided with the
LDNDC binary download and uses package fixtures derived from that example.

```{mermaid}
flowchart LR
  A["PEcAn met NetCDF"] --> B["met2model.LDNDC()"]
  B --> C["LDNDC climate input"]
  C --> D["write.config.LDNDC()"]
  D --> E["LDNDC project files"]
  E --> F["LDNDC binary"]
  F --> G["raw LDNDC output"]
  G --> H["model2netcdf.LDNDC()"]
  H --> I["yearly PEcAn NetCDF"]
```

## 1) Read the PEcAn settings

The example settings follow the same PEcAn directory pattern as Demo 1: `outdir`
contains the workflow output, `run/` contains model configuration files, and
`out/` contains model outputs.

```{r}
example_dir <- system.file("extdata", "DE_graswang", package = "PEcAn.LDNDC")
settings_path <- file.path(example_dir, "pecan.xml")

settings <- PEcAn.settings::read.settings(settings_path)
settings <- PEcAn.settings::prepare.settings(settings)

run_id <- "LDNDC-DE-GRASWANG"
dir.create(file.path(settings$rundir, run_id), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(settings$modeloutdir, run_id), recursive = TRUE, showWarnings = FALSE)

settings$host$rundir <- settings$rundir
settings$host$outdir <- settings$modeloutdir

ldndc_binary <- Sys.getenv("LDNDC_BINARY", unset = "")
if (nzchar(ldndc_binary)) {
  settings$model$binary <- ldndc_binary
}
settings$model$binary
```

If the binary is not already present at `settings$model$binary`, install it under
the PEcAn output directory and update the setting:

```{r, eval=FALSE}
ldndc_runtime <- PEcAn.LDNDC::download.LDNDC(
  location = file.path(settings$outdir, "ldndc")
)
settings$model$binary <- ldndc_runtime$binary
```

## 2) Convert meteorology with `met2model.LDNDC()`

The package fixture `ldndc.met.2019.nc` is a PEcAn-standard meteorology file.
`met2model.LDNDC()` writes the climate text file that LDNDC reads.

```{r}
met_result <- PEcAn.LDNDC::met2model.LDNDC(
  in.path = example_dir,
  in.prefix = "ldndc.met",
  outfolder = file.path(settings$rundir, run_id),
  start_date = settings$run$start.date,
  end_date = settings$run$end.date
)

settings$run$inputs$met$path <- met_result$file[[1]]
readLines(settings$run$inputs$met$path, n = 12)
```

## 3) Write LDNDC configuration with `write.config.LDNDC()`

The event, setup, site, site-parameter, and species-parameter files are fixtures
based on the grassland/meadow example bundled with LDNDC. PEcAn copies them into
the run directory and writes the generated launcher.

```{r}
settings$run$inputs$events$path1 <- file.path(example_dir, "events.xml")
settings$run$inputs$setup$path1 <- file.path(example_dir, "setup.xml")
settings$run$inputs$site$path1 <- file.path(example_dir, "site.xml")
settings$run$inputs$siteparameters$path1 <- file.path(example_dir, "siteparameters.xml")
settings$run$inputs$speciesparameters$path1 <- file.path(example_dir, "speciesparameters.xml")
settings$run$inputs$airchemistry$path1 <- file.path(example_dir, "airchemistry.txt")

trait_values <- readRDS(file.path(example_dir, "trait_values.rds"))

PEcAn.LDNDC::write.config.LDNDC(
  defaults = NULL,
  trait.values = trait_values,
  settings = settings,
  run.id = run_id
)

run_dir <- file.path(settings$rundir, run_id)
sort(list.files(run_dir))
```

The generated `project.ldndc` is the LDNDC project file, and `job.sh` is the
script PEcAn uses to run the model.

```{r}
readLines(file.path(run_dir, "project.ldndc"), n = 18)
readLines(file.path(run_dir, "job.sh"), n = 24)
```

## 4) Execute the generated run

`job.sh` launches LDNDC and calls `model2netcdf.LDNDC()` after the raw model
output is available.

```{r}
if (!file.exists(settings$model$binary)) {
  PEcAn.logger::logger.severe(
    "LDNDC binary not found at settings$model$binary. Run download.LDNDC() first."
  )
}

job_status <- system2("sh", file.path(run_dir, "job.sh"))
if (!identical(job_status, 0L)) {
  PEcAn.logger::logger.severe("LDNDC job failed; see the run logfile.")
}

out_dir <- file.path(settings$modeloutdir, run_id)
sort(list.files(out_dir))
```

## 5) Check PEcAn-standard output

The generated job normally creates the yearly NetCDF file. If raw output exists
but the NetCDF is missing, run the converter directly.

```{r}
nc_path <- file.path(out_dir, "2019.nc")
if (!file.exists(nc_path)) {
  PEcAn.LDNDC::model2netcdf.LDNDC(
    outdir = out_dir,
    sitelat = settings$run$site$lat,
    sitelon = settings$run$site$lon,
    start_date = settings$run$start.date,
    end_date = settings$run$end.date,
    delete.raw = FALSE
  )
}

file.exists(nc_path)
```

Read the PEcAn-standard output and plot a small set of variables.

```{r}
model_output <- PEcAn.utils::read.output(
  runid = run_id,
  outdir = out_dir,
  start.year = 2019,
  end.year = 2019,
  variables = c("LAI", "GPP", "NPP"),
  dataframe = TRUE,
  verbose = FALSE
)

plot(
  model_output$posix,
  model_output$LAI,
  type = "l",
  xlab = "Date",
  ylab = "LAI",
  main = "LDNDC Grassland LAI"
)
```

## Summary

The LDNDC-specific PEcAn path is:

1. `PEcAn.LDNDC::met2model.LDNDC()` writes LDNDC climate input.
2. `PEcAn.LDNDC::write.config.LDNDC()` writes the LDNDC run files.
3. The generated `job.sh` runs LDNDC.
4. `PEcAn.LDNDC::model2netcdf.LDNDC()` writes yearly PEcAn NetCDF output.

TODO: add an LDNDC-specific PEcAn events translator so the event fixture can be
generated from PEcAn event data.
