---
title: "Getting Started with LDNDC in PEcAn"
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  warning: true
  message: true
---

## Purpose

This tutorial is for **LDNDC users who have already completed Demo 1** in
`documentation/tutorials/Demo_1_Basic_Run/` and now want to learn the
LDNDC-specific PEcAn workflow.

Demo 1 covers the general PEcAn workflow. This tutorial focuses on what is
different for LDNDC:

- LDNDC is an external runtime, not just an R package
- PEcAn writes model-specific XML, text, and job files for LDNDC
- climate drivers are converted with `PEcAn.LDNDC::met2model.LDNDC()`
- raw LDNDC text outputs are converted to yearly netCDF with
  `PEcAn.LDNDC::model2netcdf.LDNDC()`

The worked example uses the currently validated grassland/meadow path and is
organized around the three LDNDC-specific coupler functions:

1. `PEcAn.LDNDC::met2model.LDNDC()`
2. `PEcAn.LDNDC::write.config.LDNDC()`
3. `PEcAn.LDNDC::model2netcdf.LDNDC()`

```{r}
ldndc_pkg_root <- here::here()
repo_root <- normalizePath(file.path(ldndc_pkg_root, "..", ".."))

run_cmd <- function(cmd, args = character(), wd = repo_root, env = character(), allow_fail = FALSE) {
  out <- withr::with_dir(
    wd,
    system2(cmd, args, stdout = TRUE, stderr = TRUE, env = env)
  )
  status <- attr(out, "status")
  if (is.null(status)) {
    status <- 0L
  }

  cat("$", paste(c(cmd, args), collapse = " "), "\n")
  if (length(out)) {
    cat(paste(out, collapse = "\n"), "\n")
  }

  if (!allow_fail && !identical(status, 0L)) {
    stop(sprintf("Command failed with status %s: %s", status, cmd), call. = FALSE)
  }

  invisible(list(status = status, output = out))
}

preview_file <- function(path, n = 12) {
  cat(paste(utils::head(readLines(path), n), collapse = "\n"), "\n")
}

write_ldndc_met_fixture <- function(target_dir) {
  source_file <- file.path(repo_root, "models", "basgra", "tests", "testthat", "test.met.2019.nc")
  target_file <- file.path(target_dir, "ldndc.met.2019.nc")
  file.copy(source_file, target_file, overwrite = TRUE)

  nc <- ncdf4::nc_open(target_file, write = TRUE)
  on.exit(ncdf4::nc_close(nc), add = TRUE)

  air_temperature <- ncdf4::ncvar_get(nc, "air_temperature")
  dims <- nc[["var"]][["air_temperature"]][["dim"]]

  nc <- ncdf4::ncvar_add(
    nc,
    ncdf4::ncvar_def("air_pressure", "Pa", dim = dims, missval = -9999)
  )
  nc <- ncdf4::ncvar_add(
    nc,
    ncdf4::ncvar_def(
      "water_vapor_saturation_deficit",
      "Pa",
      dim = dims,
      missval = -9999
    )
  )

  ncdf4::ncvar_put(
    nc,
    "air_pressure",
    array(101325, dim = dim(air_temperature))
  )
  ncdf4::ncvar_put(
    nc,
    "water_vapor_saturation_deficit",
    array(500, dim = dim(air_temperature))
  )

  target_file
}

write_tutorial_events <- function(events_path, start_date = "2019-01-01") {
  # TODO: add an LDNDC-specific PEcAn events.json -> events.txt translator
  # to the workflow and replace this tutorial helper once it exists.
  writeLines(
    c(
      "<?xml version=\"1.0\"?>",
      "<ldndcevent>",
      paste0("  <global time=\"", start_date, "\" />"),
      "  <event id=\"0\">",
      paste0("    <event type=\"plant\" time=\"", start_date, "\">"),
      "      <plant name=\"PERG\" type=\"PERG\">",
      "        <grass initialbiomass=\"500.0\" />",
      "      </plant>",
      "    </event>",
      "  </event>",
      "</ldndcevent>"
    ),
    events_path
  )
}

local_archive_name <- "ldndc-1.37.mac64.2026-03-03.tar.bz2"
public_archive_url <- paste0(
  "https://ldndc.imk-ifu.kit.edu/ldndc/downloads/public/packages/",
  "mac64/ldndc-1.37.mac64.2026-03-03.tar.bz2"
)
archive_cache_dir <- file.path(repo_root, "models", "ldndc", "vignettes", "local-assets")
dir.create(archive_cache_dir, recursive = TRUE, showWarnings = FALSE)
ldndc_archive <- file.path(archive_cache_dir, local_archive_name)

if (!file.exists(ldndc_archive)) {
  run_cmd("curl", c("-fL", public_archive_url, "-o", ldndc_archive))
}

tutorial_root <- withr::local_tempdir(pattern = "ldndc-getting-started-")
tutorial_r_lib <- file.path(tutorial_root, "Rlib")
dir.create(tutorial_r_lib)
```

## 1) Positioning and prerequisites

Before working through this tutorial, complete Demo 1:
`documentation/tutorials/Demo_1_Basic_Run/run_pecan.qmd`.

This tutorial assumes you already understand:

- what a `pecan.xml` file does
- how PEcAn settings are read and prepared
- the difference between `run/` and `out/` directories in a PEcAn run

This tutorial adds the LDNDC-specific pieces that Demo 1 does not cover.

### Local requirements

You will need:

- `R`
- `Rscript`
- `quarto`
- a local PEcAn checkout
- a user-supplied LDNDC archive cached at
  `models/ldndc/vignettes/local-assets/ldndc-1.37.mac64.2026-03-03.tar.bz2`
  or internet access to download it from the public macOS package URL used in
  this notebook

The LDNDC binary and bundled example files may be downloaded from
<https://ldndc.imk-ifu.kit.edu/download/download-model.php> and used under the
terms of the [LDNDC license](https://ldndc.imk-ifu.kit.edu/about/license.php).

```{sh}
Rscript --version
quarto --version
```

## 2) From Demo 1 to LDNDC

Demo 1 used SIPNET. For LDNDC, the overall PEcAn idea is the same, but the
model-specific pieces change:

| Demo 1 / generic PEcAn | LDNDC-specific change |
|---|---|
| `model$type = SIPNET` | `model$type = LDNDC` |
| model binary only | binary plus runtime support files |
| generic model meteorology handling | climate text written by `met2model.LDNDC()` |
| one model config file | generated `project.ldndc`, XML inputs, and `job.sh` |
| model output read directly | raw LDNDC text converted by `model2netcdf.LDNDC()` |

The `pecan.xml` shape still looks familiar. An LDNDC-oriented version of the
Demo 1 model block would look more like this:

```xml
<model>
  <type>LDNDC</type>
  <binary>/path/to/ldndc/bin/ldndc</binary>
  <delete.raw>FALSE</delete.raw>
</model>

<run>
  <site>
    <id>1</id>
    <lat>46.56</lat>
    <lon>11.03</lon>
  </site>
  <inputs>
    <met>
      <path>
        <path1>/path/to/climate.txt</path1>
      </path>
    </met>
    <events>
      <path>
        <path1>/path/to/events.xml</path1>
      </path>
    </events>
  </inputs>
  <start.date>2019/01/01</start.date>
  <end.date>2019/01/03</end.date>
</run>
```

This tutorial does not start from a full `pecan.xml`. Instead, it shows the
LDNDC-specific coupler functions directly so you can see what PEcAn needs to
generate for this model.

## 3) Set up the local tutorial workspace

We first extract the validated local runtime and install the current
`PEcAn.LDNDC` package into a temporary library.

```{r}
run_cmd("tar", c("-xjf", ldndc_archive, "-C", tutorial_root))

ldndc_root <- file.path(tutorial_root, "ldndc-1.37.mac64")
run_cmd("xattr", c("-dr", "com.apple.quarantine", ldndc_root), allow_fail = TRUE)

run_cmd(
  file.path(R.home("bin"), "R"),
  c("CMD", "INSTALL", "-l", tutorial_r_lib, file.path(repo_root, "models", "ldndc")),
  env = "TMPDIR=/tmp"
)

withr::local_libpaths(c(tutorial_r_lib, .libPaths()))
```

## 4) `met2model.LDNDC`: create LDNDC climate input

`PEcAn.LDNDC::met2model.LDNDC()` converts PEcAn meteorology into the climate text
format expected by LDNDC.

For this tutorial we use the strict fixture that was validated during the local
audit. It starts from the repository's Basgra meteorology test file and adds the
two variables LDNDC needs here: `air_pressure` and
`water_vapor_saturation_deficit`.

```{r}
met_dir <- file.path(tutorial_root, "met")
dir.create(met_dir)

met_fixture <- write_ldndc_met_fixture(met_dir)

run_root <- file.path(tutorial_root, "run")
out_root <- file.path(tutorial_root, "out")
dir.create(run_root)
dir.create(out_root)

run_id <- "TUTORIAL-RUN"
dir.create(file.path(run_root, run_id), recursive = TRUE)
dir.create(file.path(out_root, run_id), recursive = TRUE)

met_result <- PEcAn.LDNDC::met2model.LDNDC(
  in.path = met_dir,
  in.prefix = "ldndc.met",
  outfolder = file.path(run_root, run_id),
  start_date = "2019-01-01",
  end_date = "2019-01-03"
)

climate_file <- met_result$file[[1]]
climate_file
```

This is the file that will later be linked into the generated LDNDC run
directory as `climate.txt`.

```{r}
preview_file(climate_file, n = 10)
```

## 5) `write.config.LDNDC`: generate the runnable LDNDC project

`PEcAn.LDNDC::write.config.LDNDC()` writes the model-specific files PEcAn needs
to run LDNDC. For this minimal example we provide:

- the LDNDC binary path
- site latitude and longitude
- the climate file produced by `met2model.LDNDC()`
- a minimal planting event file
- a meadow PFT

```{r}
events_path <- file.path(run_root, run_id, "events.xml")
write_tutorial_events(events_path, start_date = "2019-01-01")

settings <- list(
  host = list(
    rundir = run_root,
    outdir = out_root
  ),
  rundir = run_root,
  model = list(
    binary = file.path(ldndc_root, "bin", "ldndc"),
    delete.raw = FALSE
  ),
  run = list(
    start.date = "2019-01-01",
    end.date = "2019-01-03",
    site = list(
      id = 1,
      lat = 46.56,
      lon = 11.03
    ),
    inputs = list(
      met = list(path = climate_file),
      events = list(path1 = events_path),
      airchemistry = list(path1 = NULL),
      poolinitcond = list(path = NULL),
      groundwater = list(path1 = NULL)
    )
  ),
  pfts = list(list(name = "meadow"))
)

trait_values <- list(meadow = list())

PEcAn.LDNDC::write.config.LDNDC(
  defaults = NULL,
  trait.values = trait_values,
  settings = settings,
  run.id = run_id
)

generated_rundir <- file.path(run_root, run_id)
sort(list.files(generated_rundir))
```

These generated files are the core of the LDNDC PEcAn run directory.

### `project.ldndc`

```{r}
preview_file(file.path(generated_rundir, "project.ldndc"), n = 20)
```

### `setup.xml`

```{r}
preview_file(file.path(generated_rundir, "setup.xml"), n = 20)
```

### `site.xml`

```{r}
preview_file(file.path(generated_rundir, "site.xml"), n = 16)
```

### `speciesparameters.xml`

```{r}
preview_file(file.path(generated_rundir, "speciesparameters.xml"), n = 16)
```

### `siteparameters.xml`

```{r}
preview_file(file.path(generated_rundir, "siteparameters.xml"), n = 16)
```

### `job.sh`

```{r}
preview_file(file.path(generated_rundir, "job.sh"), n = 24)
```

At this stage, the main PEcAn-to-LDNDC configuration work is done. The generated
`job.sh` is the operational bridge between PEcAn and the external LDNDC runtime.

## 6) Execute the generated run

The generated job script runs LDNDC and then calls
`PEcAn.LDNDC::model2netcdf.LDNDC()` on the raw outputs. In this tutorial we set
`R_LIBS` explicitly so the launched `Rscript` resolves the temporary package
install we just created.

```{r}
job_path <- file.path(generated_rundir, "job.sh")

run_cmd(
  "sh",
  job_path,
  env = c(
    paste0("R_LIBS=", tutorial_r_lib),
    "TMPDIR=/tmp"
  )
)

generated_outdir <- file.path(out_root, run_id)
log_file <- file.path(generated_outdir, "logfile.txt")
raw_output_dir <- file.path(generated_outdir, "Output")

sort(list.files(raw_output_dir))
```

The model log is useful when a run fails or when you want to confirm which step
completed.

```{r}
preview_file(log_file, n = 12)
```

## 7) `model2netcdf.LDNDC`: convert raw output to PEcAn netCDF

The generated job already called `model2netcdf.LDNDC()`, but it is helpful to
run the converter explicitly once so you can see its inputs and outputs.

It consumes raw text files such as:

- `physiology-subdaily.txt`
- `soilchemistry-subdaily.txt`
- `watercycle-subdaily.txt`
- `report-harvest.txt`
- `report-cut.txt`

To make the conversion step visible, we remove the existing yearly netCDF file
and recreate it directly.

```{r}
nc_path <- file.path(generated_outdir, "2019.nc")
if (file.exists(nc_path)) {
  unlink(nc_path)
}

PEcAn.LDNDC::model2netcdf.LDNDC(
  outdir = generated_outdir,
  sitelat = settings$run$site$lat,
  sitelon = settings$run$site$lon,
  start_date = settings$run$start.date,
  end_date = settings$run$end.date,
  delete.raw = FALSE
)

file.exists(nc_path)
```

You should now have a yearly netCDF file in PEcAn's expected output location.

```{r}
nc <- ncdf4::nc_open(nc_path)
on.exit(ncdf4::nc_close(nc), add = TRUE)

names(nc$var)
```

## 8) End-to-end summary

The runnable path in this tutorial is:

1. `PEcAn.LDNDC::met2model.LDNDC()` writes an LDNDC climate text file
2. `PEcAn.LDNDC::write.config.LDNDC()` writes the runnable LDNDC project files
3. the generated `job.sh` launches the LDNDC runtime
4. `PEcAn.LDNDC::model2netcdf.LDNDC()` converts raw text output to yearly netCDF

In other words:

- PEcAn prepares the model-specific inputs
- LDNDC runs externally
- PEcAn standardizes the outputs again for downstream analysis

## 9) Demo 1 bridge: how this maps back into PEcAn

If you return to a full `pecan.xml` workflow after this tutorial, the main
question is not whether PEcAn changes, but **which parts are generic PEcAn and
which parts are LDNDC-specific**.

Generic PEcAn pieces from Demo 1 that still apply:

- the overall role of `pecan.xml`
- PEcAn settings preparation and validation
- host, site, and output-directory concepts
- downstream use of standardized netCDF outputs

LDNDC-specific pieces introduced here:

- `model$type = LDNDC`
- a valid LDNDC runtime under `settings$model$binary`
- climate input generated for the LDNDC text format
- generated LDNDC XML and text inputs in the run directory
- post-processing from raw LDNDC text output to yearly netCDF

When you move from this tutorial back to a full PEcAn workflow, the easiest way
to think about it is:

- Demo 1 teaches the PEcAn workflow shell
- this tutorial teaches the LDNDC-specific pieces that fill that shell

## 10) Troubleshooting

### macOS says the LDNDC binary cannot be opened

If Gatekeeper blocks the extracted binary, remove the quarantine attribute:

```bash
xattr -dr com.apple.quarantine /path/to/ldndc-1.37.mac64
```

### `job.sh` cannot find the runtime support files

Make sure `settings$model$binary` points to the extracted LDNDC binary under a
runtime tree that also contains `dotfiles/`.

### `Rscript` cannot load `PEcAn.LDNDC`

In isolated environments, set `R_LIBS` when you launch the generated job so the
script can find the installed package library.

### The archive cache is empty

Place the LDNDC archive at:

`models/ldndc/vignettes/local-assets/ldndc-1.37.mac64.2026-03-03.tar.bz2`

or let this notebook download it from the public macOS package URL.

## Key artifacts

```{r}
cat("Archive source:", ldndc_archive, "\n")
cat("Tutorial workspace:", tutorial_root, "\n")
cat("Climate file:", climate_file, "\n")
cat("Generated run directory:", generated_rundir, "\n")
cat("Raw output directory:", raw_output_dir, "\n")
cat("NetCDF output:", nc_path, "\n")
```
