---
title: "PEcAn-LDNDC Minimal Coupler Audit Workflow"
format:
  html:
    toc: true
    code-fold: false
execute:
  echo: true
  warning: true
  message: true
---

## Purpose

This workflow performs a local execution audit of the PEcAn LDNDC coupler using:

- the local PEcAn checkout
- a single archive cache under `models/ldndc/vignettes/local-assets/`
- a public macOS package download fallback into that cache
- an isolated audit workspace under `/tmp`

It does not require `LDNDC_USER` or `LDNDC_PASS`.

```{r}
repo_root <- normalizePath(system2("git", c("rev-parse", "--show-toplevel"), stdout = TRUE))
`%||%` <- function(x, y) if (is.null(x)) y else x

run_cmd <- function(cmd, args = character(), wd = repo_root, env = character(), allow_fail = FALSE) {
  old_wd <- getwd()
  on.exit(setwd(old_wd), add = TRUE)
  setwd(wd)

  out <- system2(cmd, args, stdout = TRUE, stderr = TRUE, env = env)
  status <- attr(out, "status") %||% 0L

  cat("$", paste(c(cmd, args), collapse = " "), "\n")
  if (length(out)) {
    cat(paste(out, collapse = "\n"), "\n")
  }

  if (!allow_fail && !identical(status, 0L)) {
    stop(sprintf("Command failed with status %s: %s", status, cmd), call. = FALSE)
  }

  invisible(list(status = status, output = out))
}

local_archive_name <- "ldndc-1.37.mac64.2026-03-03.tar.bz2"
public_archive_url <- paste0(
  "https://ldndc.imk-ifu.kit.edu/ldndc/downloads/public/packages/",
  "mac64/ldndc-1.37.mac64.2026-03-03.tar.bz2"
)
archive_cache_dir <- file.path(repo_root, "models", "ldndc", "vignettes", "local-assets")
dir.create(archive_cache_dir, recursive = TRUE, showWarnings = FALSE)
ldndc_archive <- file.path(archive_cache_dir, local_archive_name)

if (!file.exists(ldndc_archive)) {
  run_cmd("curl", c("-fL", public_archive_url, "-o", ldndc_archive))
}

audit_root <- tempfile("ldndc-quarto-audit-")
dir.create(audit_root)
audit_home <- file.path(audit_root, "home")
dir.create(audit_home)
audit_r_lib <- file.path(audit_root, "Rlib")
dir.create(audit_r_lib)

replace_setting <- function(lines, key, value) {
  pattern <- paste0("^", key, "\\s*=.*$")
  replacement <- paste0(key, " = \"", value, "\"")
  if (any(grepl(pattern, lines))) {
    sub(pattern, replacement, lines)
  } else {
    c(lines, replacement)
  }
}

insert_after_setting <- function(lines, anchor_key, new_key, value) {
  new_line <- paste0(new_key, " = \"", value, "\"")
  if (any(grepl(paste0("^", new_key, "\\s*=.*$"), lines))) {
    return(replace_setting(lines, new_key, value))
  }

  anchor <- which(grepl(paste0("^", anchor_key, "\\s*=.*$"), lines))[1]
  append(lines, new_line, after = anchor)
}

write_runtime_config <- function(ldndc_root, config_path, input_path = ".", output_path = ".") {
  conf <- readLines(file.path(ldndc_root, "ldndc.conf"))
  conf <- replace_setting(conf, "resources_path", ldndc_root)
  conf <- insert_after_setting(conf, "resources_path", "udunits_path", file.path(ldndc_root, "dotfiles"))
  conf <- insert_after_setting(
    conf,
    "udunits_path",
    "service_registry_url",
    file.path(ldndc_root, "dotfiles", "service-registry.json")
  )
  conf <- insert_after_setting(conf, "service_registry_url", "input_path", input_path)
  conf <- insert_after_setting(conf, "input_path", "output_path", output_path)
  writeLines(conf, config_path)
  config_path
}

write_ldndc_met_fixture <- function(target_dir) {
  source_file <- file.path(repo_root, "models", "basgra", "tests", "testthat", "test.met.2019.nc")
  target_file <- file.path(target_dir, "ldndc.met.2019.nc")
  copied <- file.copy(source_file, target_file, overwrite = TRUE)
  if (!copied) {
    stop("Failed to copy met fixture to ", target_file, call. = FALSE)
  }

  nc <- ncdf4::nc_open(target_file, write = TRUE)
  on.exit(ncdf4::nc_close(nc), add = TRUE)

  air_temperature <- ncdf4::ncvar_get(nc, "air_temperature")
  dims <- nc[["var"]][["air_temperature"]][["dim"]]

  nc <- ncdf4::ncvar_add(
    nc,
    ncdf4::ncvar_def("air_pressure", "Pa", dim = dims, missval = -9999)
  )
  nc <- ncdf4::ncvar_add(
    nc,
    ncdf4::ncvar_def(
      "water_vapor_saturation_deficit",
      "Pa",
      dim = dims,
      missval = -9999
    )
  )

  ncdf4::ncvar_put(
    nc,
    "air_pressure",
    array(101325, dim = dim(air_temperature))
  )
  ncdf4::ncvar_put(
    nc,
    "water_vapor_saturation_deficit",
    array(500, dim = dim(air_temperature))
  )

  target_file
}
```

## 1) Environment prerequisites

```{r}
run_cmd("uname", "-a")
run_cmd("sw_vers")
run_cmd("R", "--version")
run_cmd("Rscript", "--version")
run_cmd("quarto", "--version")
```

```{r}
required <- c("ncdf4", "testthat", "readr", "lubridate", "dplyr", "pak", "withr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Missing R packages: ", paste(missing, collapse = ", "))
}
```

## 2) Extract and validate the bundled LDNDC runtime

```{r}
run_cmd("tar", c("-xjf", ldndc_archive, "-C", audit_root))

ldndc_root <- file.path(audit_root, "ldndc-1.37.mac64")
run_cmd("xattr", c("-dr", "com.apple.quarantine", ldndc_root), allow_fail = TRUE)
run_cmd("sh", "install.sh", wd = ldndc_root, env = sprintf("HOME=%s", audit_home))

direct_conf <- write_runtime_config(
  ldndc_root = ldndc_root,
  config_path = file.path(audit_root, "ldndc-direct.conf"),
  input_path = "projects",
  output_path = "projects"
)

run_cmd(
  file.path(ldndc_root, "bin", "ldndc"),
  c("-c", direct_conf, "projects/grassland/DE_graswang/DE_graswang.ldndc"),
  wd = ldndc_root
)
```

## 3) Install the local PEcAn.LDNDC package

```{r}
run_cmd(
  file.path(R.home("bin"), "R"),
  c("CMD", "INSTALL", "-l", audit_r_lib, file.path(repo_root, "models/ldndc")),
  env = "TMPDIR=/tmp"
)

.libPaths(c(audit_r_lib, .libPaths()))
```

## 4) Run package tests

```{r}
testthat::test_dir(file.path(repo_root, "models", "ldndc", "tests", "testthat"))
```

## 5) Smoke-test the coupler functions

```{r}
smoke_dir <- tempfile("ldndc-smoke-")
dir.create(smoke_dir)
dir.create(file.path(smoke_dir, "Output"))

met_dir <- tempfile("ldndc-met-")
dir.create(met_dir)
met_fixture <- write_ldndc_met_fixture(met_dir)
met_result <- PEcAn.LDNDC::met2model.LDNDC(
  met_dir,
  "ldndc.met",
  smoke_dir,
  "2019-01-01",
  "2019-01-03"
)

datetimes <- c(
  "2000-01-01 00:00:00",
  "2000-01-01 06:00:00",
  "2000-01-01 12:00:00",
  "2000-01-01 18:00:00",
  "2000-01-02 00:00:00"
)

readr::write_tsv(
  data.frame(
    datetime = datetimes,
    species = "all",
    lai = seq(1, 1.4, length.out = length(datetimes)),
    dC_co2_upt.kgCm.2. = rep(0.1, length(datetimes)),
    dC_maintenance_resp.kgCm.2. = rep(0.02, length(datetimes)),
    dC_transport_resp.kgCm.2. = rep(0.01, length(datetimes)),
    dC_growth_resp.kgCm.2. = rep(0.01, length(datetimes)),
    DW_below.kgDWm.2. = rep(0.2, length(datetimes)),
    DW_above.kgDWm.2. = rep(0.3, length(datetimes))
  ),
  file.path(smoke_dir, "Output", "physiology-subdaily.txt")
)

readr::write_tsv(
  data.frame(
    datetime = datetimes,
    sC_co2_prod_hetero.kgCm.2. = rep(0.03, length(datetimes))
  ),
  file.path(smoke_dir, "Output", "soilchemistry-subdaily.txt")
)

readr::write_tsv(
  data.frame(
    datetime = datetimes,
    soilwater_10cm... = rep(0.2, length(datetimes)),
    soilwater_30cm... = rep(0.25, length(datetimes))
  ),
  file.path(smoke_dir, "Output", "watercycle-subdaily.txt")
)

readr::write_tsv(
  data.frame(
    datetime = "2000-01-01 00:00:00",
    dC_fru_export.kgCha.1. = 0,
    dC_fol_export.kgCha.1. = 0,
    dC_frt_export.kgCha.1. = 0,
    dC_lst_above_export.kgCha.1. = 0,
    dC_lst_below_export.kgCha.1. = 0,
    dC_dst_above_export.kgCha.1. = 0,
    dC_dst_below_export.kgCha.1. = 0,
    dC_straw_export.kgCha.1. = 0
  ),
  file.path(smoke_dir, "Output", "report-harvest.txt")
)

readr::write_tsv(
  data.frame(
    datetime = "2000-01-01 00:00:00",
    dC_fru_export.kgCha.1. = 0,
    dC_fol_export.kgCha.1. = 0,
    dC_dfol_export.kgCha.1. = 0,
    dC_lst_export.kgCha.1. = 0,
    dC_dst_export.kgCha.1. = 0,
    dC_frt_export.kgCha.1. = 0
  ),
  file.path(smoke_dir, "Output", "report-cut.txt")
)

PEcAn.LDNDC::model2netcdf.LDNDC(smoke_dir, 60, 25, "2000-01-01", "2000-01-01", delete.raw = FALSE)

stopifnot(file.exists(met_fixture))
stopifnot(file.exists(met_result$file[[1]]))
stopifnot(file.exists(file.path(smoke_dir, "2000.nc")))
```

## 6) Minimal generated PEcAn run

```{r}
sample_dir <- file.path(ldndc_root, "projects", "grassland", "DE_graswang")
generated_run_root <- file.path(audit_root, "generated-rundir")
generated_out_root <- file.path(audit_root, "generated-outdir")
run_id <- "GENERATED-RUN"

dir.create(file.path(generated_run_root, run_id), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(generated_out_root, run_id), recursive = TRUE, showWarnings = FALSE)

settings <- list(
  host = list(
    rundir = generated_run_root,
    outdir = generated_out_root
  ),
  rundir = generated_run_root,
  model = list(
    binary = file.path(ldndc_root, "bin", "ldndc"),
    delete.raw = FALSE
  ),
  run = list(
    start.date = "2010-01-01",
    end.date = "2015-12-31",
    site = list(
      id = 1,
      lat = 46.56,
      lon = 11.03
    ),
    inputs = list(
      met = list(path = file.path(sample_dir, "DE_graswang_climate.txt")),
      events = list(path1 = file.path(sample_dir, "DE_graswang_event.xml")),
      airchemistry = list(path1 = NULL),
      poolinitcond = list(path = NULL),
      groundwater = list(path1 = NULL)
    )
  ),
  pfts = list(list(name = "meadow"))
)

PEcAn.LDNDC::write.config.LDNDC(
  defaults = NULL,
  trait.values = list(meadow = list()),
  settings = settings,
  run.id = run_id
)

job_path <- file.path(generated_run_root, run_id, "job.sh")
run_cmd(
  "sh",
  job_path,
  env = c(
    paste0("R_LIBS=", audit_r_lib),
    "TMPDIR=/tmp"
  )
)

generated_log <- file.path(generated_out_root, run_id, "logfile.txt")
nc_files <- Sys.glob(file.path(generated_out_root, run_id, "*.nc"))
stopifnot(length(nc_files) == 6L)
stopifnot(file.exists(generated_log))
```

## 7) Key artifacts

```{r}
cat("Archive source:", ldndc_archive, "\n")
cat("Audit workspace:", audit_root, "\n")
cat("Smoke-test netCDF:", file.path(smoke_dir, "2000.nc"), "\n")
cat("Generated-run logfile:", generated_log, "\n")
cat("Generated-run netCDF files:\n")
cat(paste(nc_files, collapse = "\n"), "\n")
```
